---
title: "two minute drill analysis"
author: "Drew Mukherjee"
date: "2022-12-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

I wanted to look at how (if it all) NFL teams perform differently in two-minute drill situations. In these spots, time is a factor and teams are in need of quick, long plays to score a touchdown or field goal before the clock expires. 

Throughout this project I'm only looking at pass plays. They can stop the clock and gain yards more quickly than runs, so they are much more relevant to two-minute drills. The drills themselves are defined as either A) two minutes or less left in the first half, or B) two minutes or less left in the second half with the team in possession being down by one score (8 points or less). This means the team with the ball will possess a sense of urgency, being within reach of tying or winning the game on that drive.

Two ways of investigating the differences are 1) comparing the outcome of two-minute drill plays to normal plays, or 2) seeing if there are patterns in which players are involved. It is possible that quarterbacks target big-time receivers more often when the game is on the line, or that more defensive attention is paid to them and the lesser-known players must step up. 

To address the first one, I'm looking at yards gained and yards after catch per completion, and location frequency of pass completions (outside, middle, short, or deep). And for the second, I'm desigining a k-mean analysis of receivers to classify them into groups, and checking the rate at which the groups are targeted in different parts of the game. My hypothesis is that pass completions in two-minute drills tend to result in more yards per play, and tend to be deep and outside (going out of bounds stops the clock). Furthermore, the prototypical WR1s (Cooper Kupp, Justin Jefferson, Davante Adams, etc.) will get more receptions during two-minute drills compared to running backs, tight ends, or slot receivers.


Importing packages

```{r packages}
library(haven)
library(tidyverse)
library(knitr)
library(ggplot2)
library(factoextra)

library(nflreadr)
library(nflfastR)

library(ClusterR)
library(cluster)
```

Data importing and cleaning to get the right format

```{r setup}
# getting play-by-play data from 2021 and 2022 (so far)
raw <- rbind(load_pbp(2021), load_pbp(2022))
raw$posteam_mov <- raw$posteam_score - raw$defteam_score

# selecting important variables
sel <- c('passer', 'passer_id', 'receiver', 'receiver_id', 'posteam', 'complete_pass', 'incomplete_pass', 
         'interception', 'fumble', 'sack', 'qb_scramble', 'pass_length',
         'pass_location', 'yards_gained','air_yards', 'yards_after_catch', 'shotgun', 
         'no_huddle', 'home_team','away_team','posteam_type','yardline_100',
         'posteam_score', 'defteam_score', 'posteam_mov', 'qtr', 'half_seconds_remaining', 
         'game_seconds_remaining','down','ydstogo','posteam_timeouts_remaining',
         'defteam_timeouts_remaining', 'qb_dropback')

# filtering for all QB dropbacks
dropback <- raw[,..sel] %>% filter(qb_dropback == 1)
rm(raw)
dropback$outside_deep <- ifelse(dropback$pass_length == "deep" & 
                                  (dropback$pass_location == "right" | dropback$pass_location == "left"), 1, 0)
dropback$outside_short <- ifelse(dropback$pass_length == "short" & 
                                  (dropback$pass_location == "right" | dropback$pass_location == "left"), 1, 0)
dropback$middle_deep <- ifelse(dropback$pass_length == "deep" & dropback$pass_location == "middle", 1, 0)
dropback$middle_short <- ifelse(dropback$pass_length == "short" & dropback$pass_location == "middle", 1, 0)


# checking for min pass attempts/receptions
pass_attempts <- aggregate(qb_dropback ~ passer + passer_id, data = dropback, FUN=length) %>% filter(qb_dropback >= 100)
receptions <- aggregate(qb_dropback ~ receiver + receiver_id, data = dropback, FUN=length) %>% filter(qb_dropback >= 50)

# calculate stats for players if they pass the min threshold
total_qb_stats <- merge(pass_attempts,
                      aggregate(cbind(complete_pass, incomplete_pass, shotgun, 
                                      air_yards, yards_gained) 
                                ~ passer + passer_id, data = dropback, FUN=sum, 
                                na.rm=TRUE))
total_rec_stats <- merge(receptions, aggregate(cbind(yards_after_catch, yards_gained, 
                                                     outside_deep, outside_short, 
                                                     middle_deep, middle_short) 
                                 ~ receiver + receiver_id, data = dropback, FUN=sum, 
                                 na.rm=TRUE))

avg_qb_stats <- merge(pass_attempts,
                        aggregate(cbind(complete_pass, incomplete_pass, shotgun, air_yards, yards_gained) 
                                  ~ passer + passer_id, data = dropback, FUN=mean, na.rm=TRUE))
avg_rec_stats <- merge(receptions, 
                         aggregate(cbind(yards_after_catch, yards_gained, outside_deep, 
                                         outside_short, middle_deep, middle_short) 
                                   ~ receiver + receiver_id, data = dropback, FUN=mean, na.rm=TRUE))
rm(pass_attempts); rm(receptions)

avg_rec_stats <- avg_rec_stats[order(-avg_rec_stats$qb_dropback),] %>% mutate_if(is.numeric, ~round(., 3))
total_rec_stats <- total_rec_stats[order(-total_rec_stats$qb_dropback),] %>% mutate_if(is.numeric, ~round(., 3))

# nifty trick, the only players that have average YAC > average yards gained are running backs (and Deebo)
rbs <- avg_rec_stats %>% filter(yards_after_catch > yards_gained)
wr_te <- avg_rec_stats %>% filter(avg_rec_stats$yards_after_catch <= avg_rec_stats$yards_gained)

#qb_vars
rec_vars <- c('yards_after_catch', 'yards_gained',  'outside_deep', 'outside_short', 
              'middle_deep', 'middle_short')

scaled_rec <- data.frame(scale(avg_rec_stats[,rec_vars]))
scaled_rec <- apply(scaled_rec, 2, function(x) as.numeric(x))
#scaled_rec <- round(scaled_rec, 3)

# finding optimal number of clusters
fviz_nbclust(scaled_rec, kmeans, method = "wss", k.max = 8)

#k4 <- kmeans(scaled_rec, centers = 4, nstart = 25)
#fviz_cluster(k4, data=scaled_rec)

k3 <- kmeans(scaled_rec, centers=3, nstart = 25)
fviz_cluster(k3, data=scaled_rec)

#scaled_rec <- cbind(total_rec_stats$receiver, scaled_rec, k4$cluster)
total_rec_stats <- cbind(total_rec_stats, k3$cluster)
avg_rec_stats <- cbind(avg_rec_stats, k3$cluster)

total_rec_stats[which(total_rec_stats$receiver == 'C.McCaffrey'),ncol(total_rec_stats)]
total_rec_stats[which(total_rec_stats$receiver == 'C.Kupp'),ncol(total_rec_stats)]
total_rec_stats[which(total_rec_stats$receiver == 'A.Thielen'),ncol(total_rec_stats)]
total_rec_stats[which(total_rec_stats$receiver == 'T.Kelce'),ncol(total_rec_stats)]

# finding summary stats for each category
cat1 <- aggregate(cbind(yards_after_catch, yards_gained) ~ k3$cluster, 
                              data = avg_rec_stats, FUN=mean)
categories <- round(merge(cat1, aggregate(cbind(outside_deep, outside_short, 
                                                middle_deep, middle_short) 
                                          ~ k3$cluster, data = total_rec_stats, FUN=sum)),2)

categories[nrow(categories)+1,] <- NA
categories$total <- NA
for (j in 2:ncol(categories)) {
  categories[nrow(categories),j] <- sum(categories[1:nrow(categories)-1,j])
}
for (i in 1:nrow(categories)) {
  categories$total[i] <- sum(categories[i,4:7]) # shouldn't really be hardcoded
}

categories[,4:7] <- round((categories[,4:7] / categories$total),3)
```

# Receiver Categories

**Cluster 1: 7.09 YAC, 8.02 avg. yards, 98% of receptions in the outside
of the field**

-   YAC masters, low yards per reception, mostly RBs catching checkdowns and low-yardage TEs/slot WRs

-   Christian McCaffrey, Cole Beasley, Joe Mixon, Saquon Barkley, Tyler Higbee

**Cluster 2: 20% of receptions are deep (avg is 12%), 4.07 YAC, 13.61 avg
yards, mostly outside**

-   Prototypical wide receivers, lots of deep shots (mostly outside), most efficient receivers

-   Cooper Kupp, Davante Adams, Ja'Marr Chase, Justin Jefferson, Stefon
    Diggs

**Cluster 3: 28% of receptions are in the middle (avg 22%), mostly
short, 4.49 YAC, 11.14 avg yards**

-   Reliable, consistent players to get short yardage, big targets over the middle, redzone threats

-   Chris Godwin, George Kittle, Mark Andrews, Michael Pittman, Travis
    Kelce
    
    
```{r twomin}

# finding tendency differences for two-minute drills
dropback <- inner_join(dropback, avg_rec_stats[,c('receiver_id', 'k3$cluster')], by='receiver_id')

# just dropbacks during two minute drills in one-score games
two_min <- dropback %>% filter((half_seconds_remaining <= 120 & qtr == 2) |
                               (half_seconds_remaining <= 120 & qtr == 4 & 
                                  posteam_mov <= 0 & posteam_mov >= -8))

twomin_tends <- summarize(two_min, 
                          yards_after_catch = mean(yards_after_catch, na.rm=TRUE),
                          yards_gained = mean(yards_gained, na.rm=TRUE),
                          outside_deep = mean(outside_deep, na.rm=TRUE),
                          outside_short = mean(outside_short, na.rm=TRUE),
                          middle_deep = mean(middle_deep, na.rm=TRUE),
                          middle_short = mean(middle_short, na.rm=TRUE),
                          cluster1 = (length(which(`k3$cluster` == 1))) / length(`k3$cluster`),
                          cluster2 = (length(which(`k3$cluster` == 2))) / length(`k3$cluster`),
                          cluster3 = (length(which(`k3$cluster` == 3))) / length(`k3$cluster`))

all_tends <- round(rbind(twomin_tends, 
                         summarize(dropback, 
                                   yards_after_catch = mean(yards_after_catch, na.rm=TRUE),
                                   yards_gained = mean(yards_gained, na.rm=TRUE),
                                   outside_deep = mean(outside_deep, na.rm=TRUE),
                                   outside_short = mean(outside_short, na.rm=TRUE),
                                   middle_deep = mean(middle_deep, na.rm=TRUE),
                                   middle_short = mean(middle_short, na.rm=TRUE),
                                   cluster1 = (length(which(`k3$cluster` == 1))) / length(`k3$cluster`),
                                   cluster2 = (length(which(`k3$cluster` == 2))) / length(`k3$cluster`),
                                   cluster3 = (length(which(`k3$cluster` == 3))) / length(`k3$cluster`))), 3)

rownames(all_tends) <- c('Two Minute Drills', 'All Plays')
```

# Two-Minute Drill Tendencies

Overall, there aren't any clear differences in teams' tendencies when they're in two-minute drills. Receivers have slightly more deep outside receptions than normal (15.4% compared to 14.6%) but fewer short outside receptions (60.3% to 63.1%), so QBs aren't necessarily aiming for the sidelines. Each cluster has a nearly identical proportion of catches (within 2 percentage points for each), perhaps surprising since common sense would dictate the 'big play' WR1s would be more impactful during this time.

Looking into targets instead of just catches would probably lend some more insight, as well as separating by QB, offensive scheme, etc. But the initial steps of getting the data set up and being able to classify WRs in a way that aligns with the eye test signify a good start. 
